Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-02-16 submission/nlp_models.tex
some typos/tweaks, a few more to come

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
%
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can indentify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously,
or what we call a series of ``good-better-best'' models.

The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{huggingface}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrics.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
%
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
%
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12, 24, 36, \text{and } 48$ layers, respectively, with increasingly larger weight matrices.
%The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.

@@

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unsually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$, and it also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the ``bad'' GPT versus the ``good'' GPT2-small.
%
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

@@

We next describe the behavior of the Spectral norm in GPT versus GPT2-small.
In Figure~\ref{fig:GPT-snorm-hist}, the ``bad'' GPT model has a smaller mean/median Spectral norm as well as, spuriously, many much smaller Spectral norms, compared to the ``good'' GPT2-small, violating the conventional wisdom that smaller Spectral norms are better.
Indeed, because there are so many anonymously small Spectral norms, it appears that the GPT model may be exhibiting a kind of \emph{Scale Collapse}, like that observed in the distilled CV models (in Figure~\ref{fig:resnet204D5L}).
This is important because it demonstrates that, while the Spectral norm may correlate well with predicted test error, it is \emph{not} a good indicator of the overall model quality.
It can mispredict good-versus-bad questions in ways not seen with PL-based metrics.
Using it as an empirical quality metric may give spurious results when applied to poorly-trained or otherwise deficient models.

@@

We also examine in Figure~\ref{fig:gpt-alpha-layers} the PL exponent $\alpha$ and Log Spectral Norm versus layer id, for GPT and GPT2-small.
Let's start with Figure~\ref{fig:gpt-alpha-layer}, which plots $\alpha$ versus the depth (i.e., layer id) for each model.
The deficient GPT model displays two trends in $\alpha$, one stable with $\alpha\sim 4$, and one increasing with layer id, with $\alpha$ reaching as high as $12$.
In contrast, the well-trained GPT2-small model shows consistant and stable patterns, again with one stable $\alpha\sim 3.5$ (and below the GPT trend), and the other only slightly trending up, with $\alpha\le 6$. 
The scale-invariant $\alpha$ metric lets us identify potentially poorly-trained models; and these results show that the Correlation Flow differs significantly between GPT and GPT2-small (with the better GPT2-small looking more like the better ResNet-1K from Figure~\ref{fig:resnet-alpha-layer}).

@@

\begin{figure}[htb]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-alpha-depth.png}
        \includegraphics[width=5.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    %KDD% 
    \quad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-snorm-depth.png}
        \includegraphics[width=5.5cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{PL exponents ($\alpha$) (in (a)) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) (in (b)) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.  (Note that the Y axes on each plot are different.)
             In the text, this is interpreted in terms of \emph{Correlation Flow} and \emph{Scale~Collapse}.
            }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

We now look across series of increasingly improving GPT2 models (i.e., we consider good-better-best questions), by examining both the PL exponent $\alpha$ as well as the Log Norm metrics.  
In general, as we move from GPT2-medium to GPT2-xl, histograms for both $\alpha$ exponents and the Log Norm metrics downshift from larger to smaller values. 
For example, see Figure \ref{fig:gpt2-histograms}, which shows the histograms over the layer weight matrics for fitted PL exponent ($\alpha$) and the Log Alpha Norm ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$)~metric.

>>>>

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
%
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can indentify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously,
an example of what we call a series of ``good-better-best'' models.

The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{huggingface}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrics.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
%
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
%
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12$, $24$, $36$, and $48$ layers, respectively, with increasingly larger weight matrices.
%The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.

@@

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unsually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$.
It also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the ``bad'' GPT versus the ``good'' GPT2-small.
%
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

@@

We next describe the behavior of the Spectral norm in GPT versus GPT2-small.
In Figure~\ref{fig:GPT-snorm-hist}, the ``bad'' GPT model has a smaller mean/median Spectral norm as well as, spuriously, many much smaller Spectral norms, compared to the ``good'' GPT2-small, violating the conventional wisdom that smaller Spectral norms are better.
Indeed, because there are so many anonymously small Spectral norms, it appears that the GPT model may be exhibiting a kind of \emph{Scale Collapse}, like that observed in the distilled CV models (in Figure~\ref{fig:resnet204D5L}).
This is important because it demonstrates that, while the Spectral (or Frobenius) norm may correlate well with predicted test error, it is \emph{not} a good indicator of the overall model quality.
It can mispredict good-versus-bad questions in ways not seen with PL-based metrics.
Using it as an empirical quality metric may give spurious results when applied to poorly-trained or otherwise deficient models.

@@

We also examine in Figure~\ref{fig:gpt-alpha-layers} the PL exponent $\alpha$ and Log Spectral Norm versus layer id, for GPT and GPT2-small.
Let's start with Figure~\ref{fig:gpt-alpha-layer}, which plots $\alpha$ versus the depth (i.e., layer id) for each model.
The deficient GPT model displays two trends in $\alpha$, one stable with $\alpha\sim 4$, and one increasing with layer id, with $\alpha$ reaching as high as $12$.
In contrast, the well-trained GPT2-small model shows consistant and stable patterns, again with one stable $\alpha\sim 3.5$ (and below the GPT trend), and the other only slightly trending up, with $\alpha\le 6$. 
The scale-invariant $\alpha$ metric lets us identify potentially poorly-trained models.
These results show that the Correlation Flow differs significantly between GPT and GPT2-small (with the better GPT2-small looking more like the better ResNet-1K from Figure~\ref{fig:resnet-alpha-layer}).

@@

\begin{figure}[htb]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-alpha-depth.png}
        \includegraphics[width=5.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    %KDD% 
    \quad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-snorm-depth.png}
        \includegraphics[width=5.5cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{PL exponents ($\alpha$) (in (a)) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) (in (b)) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.  (Note that the quantities being shown on each Y axis are different.)
             In the text, this is interpreted in terms of \emph{Correlation Flow} and \emph{Scale~Collapse}.
            }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

We now look across series of increasingly improving GPT2 models (i.e., we consider good-better-best questions), by examining both the PL exponent $\alpha$ as well as the Log Norm metrics.  
In general, as we move from GPT2-medium to GPT2-xl, histograms for both $\alpha$ exponents and the Log Norm metrics downshift from larger to smaller values. 
For example, see Figure \ref{fig:gpt2-histograms}, which shows the histograms over the layer weight matrices for fitted PL exponent ($\alpha$) and the Log Alpha Norm ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$)~metric.
